Unshuffling Permutations

### 2015-07-01 article.tex
Partie algebre (suite ; en plein dans les modifs)

\begin{proposition} \label{prop:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_{\widetilde{\,}}, \: \phi_{\bar \,}, \: \phi_{-1}
        : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, or $\pi^{-1}$ are endomorphims of associative
    algebras.
\end{proposition}
\begin{proof}
    ADAPT NEXT ARGUMENTS AND FINISH.

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\rho}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$
        is one of its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$
        is one of its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
    Since $\pi$ is a square permutation and $\rho$ is a square root
    of $\pi$, by definition $\pi$ appears in the product
    $\rho \SHUFFLE \rho$. Now, by Proposition~\ref{prop:endomorphisms},
    since $\phi_{\widetilde{\,}}$ is a morphism of associative algebras
    from $\QQ[S]$ to $\QQ[S]$, $\phi_{\widetilde{\,}}$ commutes with
    the shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_{\widetilde{\,}}(\rho \SHUFFLE \rho) =
        \phi_{\widetilde{\,}}(\rho) \SHUFFLE \phi_{\widetilde{\,}}(\rho).
    \end{equation}
    Then, since $\pi$ appears in $\rho \SHUFFLE \rho$,
    $\widetilde{\pi}$ appears in $\widetilde{\rho} \SHUFFLE \widetilde{\rho}$.
    This implies~\ref{item:square_stability_1}.
    \smallskip

    The proofs of~\ref{item:square_stability_2}
    and~\ref{item:square_stability_3} are analogous.
\end{proof}
\medskip

>>>>

\begin{proposition} \label{prop:endomorphisms}
    The three linear maps
    \begin{equation}
        \phi_1, \phi_2, \phi_3 : \QQ[S] \to \QQ[S]
    \end{equation}
    linearly sending a permutation $\pi$ to, respectively,
    $\widetilde{\pi}$, $\bar \pi$, or $\pi^{-1}$ are endomorphims of
    associative algebras.
\end{proposition}
\begin{proof}
    To prove that, for $j = 1, 2, 3$, $\phi_j$ is a morphism of
    associative algebras, we have to prove that for all permutations
    $\sigma$ and $\nu$,
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \nu) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\nu).
    \end{equation}
    We shall proceed here as we have done during the proof of
    Proposition~\ref{prop:shuffle_associative_commutative} by showing
    that $\phi_j$ is a morphism of coalgebras, that is
    \begin{equation}
        \Delta \phi_j = (\phi_j \otimes \phi_j) \Delta.
    \end{equation}
    \smallskip

    Let $\pi$ be a permutation. If $P$ is a set of indexes of letters of
    $\pi$, we denote by $\widetilde{\pi}$ the set
    $\{|\pi| - i  + 1 : i \in P_1\}$. Now since this operation
    $\widetilde{\,}$ define a bijection on the set of the subsets of
    $[|\pi|]$ and the standardisation operation commutes with the
    mirror operation on words, we have
    \begin{equation} \begin{split}
        \Delta(\phi_1(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]} \STD\left(\phi_1(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_1(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 =
        [|\pi|]} \STD\left(\widetilde{\pi}_{|\widetilde{P_1}}\right)
        \otimes \STD\left(\widetilde{\pi}_{|\widetilde{P_2}}\right) \\
        & = \sum_{P_1 \uplus P_2 =
        [|\pi|]} \STD\left(\widetilde{\pi}_{|P_1}\right)
        \otimes \STD\left(\widetilde{\pi}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 =
        [|\pi|]} \phi_1\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_1\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_1 \otimes \phi_1) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_1$ is a morphism of coalgebras and hence, that
    $\phi_1$ is a morphism of associative algebras.
    \smallskip

    ADAPT NEXT ARGUMENTS AND FINISH.

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\rho}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    direct consequences of Proposition~\ref{prop:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\rho$ is a square root of
    $\pi$, by definition $\pi$ appears in the product
    $\rho \SHUFFLE \rho$. Now, by Proposition~\ref{prop:endomorphisms},
    for all $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\rho \SHUFFLE \rho) =
        \phi_j(\rho) \SHUFFLE \phi_j(\rho).
    \end{equation}
    Then, since $\pi$ appears in $\rho \SHUFFLE \rho$, $\phi_j(\pi)$
    appears in $\phi_j(\rho) \SHUFFLE \phi_j(\rho)$. This
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
\end{proof}
\medskip
